Approximation as Differentiation

### 2025-06-19 main/first-order.tex

For a first-order language with a primitive type $\syn{nat}$ and primitive operations such as $+$, the
language would provide a model of $\syn{nat}$ and its operations in $\Fam(\LatGal)$. For example, we might
interpret $\syn{nat}$ to the object $(\Nat,n \mapsto \downset{n})$, where the family component assigns to
every natural number $n$ the two-point lattice $\downset{n} = \{n, \bot\}$ indicating whether $n$ is fully
present ($n$) or fully absent ($\bot$) from an approximation of $n$, as shown in the example in
\secref{introduction}. A slight variant is to use the constant family which assigns to every $n$ the two-point
lattice $\Two = \{\mathsf{tt},\mathsf{ff}\}$; using $\downset{n}$ is slightly more ergonomic as the
approximants of a given value then contain, rather than merely determining, all the information about the
value that hasn't been approximated away.

We might then interpret $+$ as the morphism $(+,\partial+)$, i.e.~the usual addition operator + paired with an
approximating version $\partial+_{n,m}: \downset{n} \to \downset{m}$ defined as:
\begin{displaymath}
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \partial+_{n,m}&n&m&=&n + m \\
      \partial+_{n,m}&\bot&\_&=&\bot \\
      \partial+_{n,m}&\_&\bot&=&\bot
    \end{array}
\end{displaymath}
These elementary approximation structures would then be combined by the interpretation of composite types like
sums and products, providing suitable derived structures for composite values.

@@

A programmer, or perhaps a language implementor, might also like to embed additional ``approximation points''
into composite structures, to allow for example the list $1 : 2 : []$ to have the partial list $1 : \bot$ as
an approximant, representing a list with at least one element. This can be accomplished by extending the
syntax with a \emph{lifting} type constructor $\synVar{L}$. Then, as an alternative to the usual definition of
lists of natural numbers as $\mu\alpha.1 + \synVar{nat} \times \alpha$, one could (within the language) define
a type of ``approximation lists'' $\mu\alpha.\synVar{L}\,(1 + \synVar{L}\,\synVar{nat} \times \alpha)$, with
monadic computations over approximated lists computing partial lists like $1 : \bot$. Similarly, one could
define a custom approximating version of multiplication $*_{\text{appr}}: \synVar{L}\,\synVar{nat} \times
\synVar{L}\,\synVar{nat} \to \synVar{L}\,\synVar{nat}$, as:
\[x *_{\text{appr}} y = \tmBind{x}{\tmFun{x'}{\syn{if}\;x' \equiv
0\;\syn{then}\;\tmReturn{0}\;\syn{else}\;\tmBind{y}{\tmFun{y'}{\tmReturn{x' * y'}}}}}\]

@@

With this lifting operator in the core language, one could then choose to fix certain choices of
approximation, by providing a surface language without the lifting monad and interpreting it into the core
language with lifting. The Galois slicing implementations discussed in \citet{perera12a} and
\citet{ricciotti17}, for example, provide an approximation point at every composite type constructor; we can
emulate this in our approach by defining a surface language where the data type of lists $\mu\alpha.1 +
\synVar{nat} \times \alpha$ would desugar automatically into the approximating version
$\mu\alpha.\synVar{L}\,(1 + \synVar{L}\,\synVar{nat} \times \alpha)$.

To interpret the syntactic type constructor $\synVar{L}$, consider the strong monad $(\mathcal{L}, \eta, \mu)$
which extends any poset $X$ with a distinguished bottom element $\bot$ and any monotone function $f$ between
posets with a mapping from $\bot$ to $\bot$. $\mathcal{L}$ is also a strong monad on meet-semilattices. On
(bounded) join-semilattices, $\mathcal{L}$ induces a costrong comonad $(\mathcal{L}, \varepsilon, \delta)$,
with the bottom element of $X$ needed to implement the counit $\varepsilon_X: \mathcal{L}(X) \to X$. These two
combine to form a strong lifting monad in $\LatGal$ with unit $\varepsilon_X \dashv \eta_X$ and multiplication
$\delta_X \dashv \mu_X$.

>>>>

For a first-order language with a primitive type $\syn{nat}$ and primitive operations such as $+$, the
language would provide a model of $\syn{nat}$ and its operations in $\Fam(\LatGal)$. For example, we might
interpret $\syn{nat}$ as the object $(\Nat,n \mapsto \downset{n})$, where the family component assigns to
every natural number $n$ the two-point lattice $\downset{n} = \{n, \bot\}$ indicating whether $n$ is fully
present ($n$) or fully absent ($\bot$) from an approximation of $n$, as shown in the example in
\secref{introduction}. A slight variant is to use the constant family which assigns to every $n$ the two-point
lattice $\Two = \{\mathsf{tt},\mathsf{ff}\}$; using $\downset{n}$ can be more useful in practice as the
approximants of a given value then contain all the information about the value that hasn't been approximated
away.

We might interpret $+$ as the morphism $(+,\partial+)$, i.e.~the usual addition operator + paired with an
approximating version $\partial+_{n,m}: \downset{n} \to \downset{m}$ defined as:
\begin{displaymath}
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \partial+_{n,m}&n&m&=&n + m \\
      \partial+_{n,m}&\bot&\_&=&\bot \\
      \partial+_{n,m}&\_&\bot&=&\bot
    \end{array}
\end{displaymath}
These elementary approximation structures would then be combined by the interpretation of composite types like
sums and products, providing suitable derived structures for composite values.

@@

A programmer, or perhaps a language implementor, might want to embed additional ``approximation points'' into
composite data types, to allow for richer approximation structures. For example, for a list like $1 : 2 : []$
we might want to allow the partial lists $\bot$, $1 : \bot$ and $1 : 2 : \bot$ as approximants, in addition to
the approximants obtained by replacing elements with $\bot$. This can be accomplished by extending the syntax
with a \emph{lifting} type constructor $\synVar{L}$. Then, as an alternative to the usual definition of lists
of natural numbers as $\mu\alpha.1 + \synVar{nat} \times \alpha$, one could (within the language) define a
type of ``approximation lists'' $\mu\alpha.\synVar{L}\,(1 + \synVar{L}\,\synVar{nat} \times \alpha)$, with
monadic computations over approximated lists computing partial lists like $1 : \bot$. Similarly, one could
define a custom approximating version of multiplication $*_{\text{appr}}: \synVar{L}\,\synVar{nat} \times
\synVar{L}\,\synVar{nat} \to \synVar{L}\,\synVar{nat}$, as:
\[x *_{\text{appr}} y = \tmBind{x}{\tmFun{x'}{\syn{if}\;x' \equiv
0\;\syn{then}\;\tmReturn{0}\;\syn{else}\;\tmBind{y}{\tmFun{y'}{\tmReturn{x' * y'}}}}}\]

@@

With this lifting operator in the core language, one could then choose to fix certain choices of
approximation, by providing a surface language without the lifting monad and interpreting it into the core
language with lifting. The Galois slicing implementations discussed in \citet{perera12a} and
\citet{ricciotti17}, for example, provide an approximation point at every composite type constructor; we can
emulate this in our approach by defining a surface language where the data type of lists $\mu\alpha.1 +
\synVar{nat} \times \alpha$ desugars automatically into the approximating version $\mu\alpha.\synVar{L}\,(1 +
\synVar{L}\,\synVar{nat} \times \alpha)$.

To interpret the syntactic type constructor $\synVar{L}$, consider the strong monad $(\mathcal{L}, \eta, \mu)$
which acts on posets $X$ to extend them with a distinguished bottom element $\bot$ and which acts on monotone
functions $f$ between posets to extend them with a mapping from $\bot$ to $\bot$. $\mathcal{L}$ is also a
strong monad on meet-semilattices. On (bounded) join-semilattices, $\mathcal{L}$ induces a costrong comonad
$(\mathcal{L}, \varepsilon, \delta)$, with the bottom element of $X$ needed to implement the counit
$\varepsilon_X: \mathcal{L}(X) \to X$. These two combine to form a strong lifting monad in $\LatGal$ with unit
$\varepsilon_X \dashv \eta_X$ and multiplication $\delta_X \dashv \mu_X$.

### 2025-06-19 main/introduction.tex

\section{Introduction}

To audit any computational process, we need robust and well-founded notions of \emph{provenance} to track how data are used. This allows us to answer questions like ``Where did these data come from?'', ``Why are these data in the output?'' and ``How were these data computed?''. Provenance tracking has a wide range of applications, from debugging and program comprehension \todo{cite1} to improving reproducibility and transparency in scientific workflows \todo{cite2}. \emph{Program slicing}, first proposed by~\citet{weiser81}, is a collection of techniques for provenance tracking that attempts to take a run of a program and areas of interest in the output, and turn them into the subset of the input and the program that were responsible for generating those specific outputs.

>>>>

\section{Introduction}
\label{sec:introduction}

To audit any computational process, we need robust and well-founded notions of \emph{provenance} to track how data are used. This allows us to answer questions like ``Where did these data come from?'', ``Why are these data in the output?'' and ``How were these data computed?''. Provenance tracking has a wide range of applications, from debugging and program comprehension~\cite{buneman95,cheney07} to improving reproducibility and transparency in scientific workflows~\cite{kontogiannis08}. \emph{Program slicing}, first proposed by~\citet{weiser81}, is a collection of techniques for provenance tracking that attempts to take a run of a program and areas of interest in the output, and turn them into the subset of the input and the program that were responsible for generating those specific outputs.

### 2025-06-19 notes/stability.tex
[ tex ] rough attempt to develop the stablity<->differentiable analogy

\newcommand{\downset}[1]{\mathop{\downarrow}(#1)}

@@

Stability was invented by Berry \cite{berry86} in an attempt to capture
sequentiality, and were also instrumental in the discovery of Linear
Logic via Coherence Spaces \cite{girard}. A textbook description of
stable functions in the context of domain theory is given by Amadio
and Curien \cite{amadio-curien} (Chapter 12). Here, we are not
concerned with completeness properties considered in domain theory, so
we start with only partially ordered sets. Stable functions can be
defined as a certain subset of the set of monotone functions between
two partially ordered sets:

>>>>

@@

Stability was invented by Berry \cite{berry79} in an attempt to capture
sequentiality, and were also instrumental in the discovery of Linear
Logic via Coherence Spaces \cite{girard}. A textbook description of
stable functions in the context of domain theory is given by Amadio
and Curien \cite{amadio-curien} (Chapter 12). Here, we are not
concerned with completeness properties considered in domain theory, so
we start with only partially ordered sets. Stable functions can be
defined as a certain subset of the set of monotone functions between
two partially ordered sets:
